---
title: "Data types and structures"
author: Ralf Ansorg
date: "`r Sys.Date()`"
format:
  typst:
    template-partials: 
      - typst-show.typ
      - typst-template.typ
echo: true
results: true
warning: false
---

## Exercise 1

Suppose you got the following data or data frame:

```{r}
#| label: init
a <- c(1:5) # mouse ID
b <- c(5, 6, 3, 10, 7) # licks per hours
c <- c("17.18", "16.03", "15.9", "17.99", "14") # length in cm
d <- c("Female", "Male", "Male", "Female", "Maile") # sex of mouse
e <- c(TRUE, FALSE, TRUE, TRUE, FALSE) # healthy mouse
```

```{r}
#| label: init-df
df <- data.frame(a=a, b=b, c=c, d=d, e=e)
```

1.  Which vectors (or columns of the data frame) need some work-over? If so, please write the correct code (_hint_: try to figure this out by looking at its structure; recall coercion and accessing elements of vectors, and factors. Think about on how to assign new a value to a particular element of a vector, e.g. my_example_vector[5].

```{r}
#| label: ex1-1a
class(c)
c
```

```{r}
#| label: ex1-1b
c <- as.numeric(c)
c
class(c)
```

```{r}
#| label: ex1-1c
d[5] <- "Male" # was misspelled as "Maile"
d <- as.factor(d)
d
```

2.  Add all vectors to a new data frame `df_clean` with proper/meaningful headers (instead of a,..., e)

```{r}
#| label: ex1-2
df_clean <- data.frame(ID=a, licks=b, length=c, sex=d, healthy=e)
df_clean
```

3.  What is the mean of the length of the mice? _Hint_: you could use the `mean()` function. What is the median?

```{r}
#| label: ex1-3
mean(df_clean$length)
median(df_clean$length)
```

```{r}
#| label: ex1-3b
#| echo: false
#| results: false
mean <- mean(df_clean$length)
median <- median(df_clean$length)
```

> The mean length of the mice is `r mean` and the median is `r median`.[^1]

4.  Look at the structure **and** summary if your cleaned-up data is reasonable and briefly explain why. Bullet points only (use the \* or \- symbol), no essay.

```{r}
#| label: ex1-4
str(df)
summary(df)

str(df_clean)
summary(df_clean)
```

## Comments

* length had to be converted to numerical (from character)
* sex should be a factor
* factors and logicals are displayed differently (counts)

[^1]: Check the .qmd file on how to have inline code!